---
title: "Cyclones Introduction"
author: "Ben Branoff"
date: "2025-03-18"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

R utilities for modeling and analyzing meteorological cyclones. Fit models to observations of storm characteristics and use these models to predict and interpolate wind and pressure fields across a storm's area. 


<details>
<summary>Read in and parse previous cyclone data</summary>
Start with the IBTrACS data (updated data available at: https://www.ncei.noaa.gov/data/international-best-track-archive-for-climate-stewardship-ibtracs/v04r01/access/csv/)

```{r loading, include=TRUE,warning=FALSE,message=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(purrr)

###  Load in the North Atlantic data only:
IBTrACS <- read.csv(paste0(dirname(getwd()),"./data/ibtracs.NA.list.v04r01.csv"))
####  limit to storms in or after 2000 for brevity
IBTrACS <- IBTrACS %>%
  filter(SEASON>=2000)
```

Determine the extent of maximum winds and other wind extent observations from the data
```{r extract data, include=TRUE,warning=FALSE,message=FALSE}
quaddist <- bind_rows(
  IBTrACS %>%  
    ####  these columns describe the extent of 34,50 and 64 knot winds in each of the storms quadrant
    ####  at different timesteps 
    select(USA_R34_NE,USA_R34_SE,USA_R34_NW,USA_R34_SW,  
           USA_R50_NE,USA_R50_SE,USA_R50_NW,USA_R50_SW,  
           USA_R64_NE,USA_R64_SE,USA_R34_NW,USA_R64_SW,
           ###  these are identifiers and more variables important for additional models
           SID,ISO_TIME,USA_SSHS,USA_EYE,USA_RMW,USA_WIND,USA_PRES,USA_ROCI,USA_POCI,USA_PRES) %>%
    tidyr::pivot_longer(cols=c(USA_R34_NE,USA_R34_SE,USA_R34_NW,USA_R34_SW,  
                               USA_R50_NE,USA_R50_SE,USA_R50_NW,USA_R50_SW,  
                               USA_R64_NE,USA_R64_SE,USA_R34_NW,USA_R64_SW),  
                        names_to =c("speed","quad"),  
                        names_pattern = "USA_R(.*)_(.*)",values_to="dist") %>%  
    mutate(speed=as.numeric(speed),  
           source="swaths"),  
  IBTrACS %>%  
    select(SID,ISO_TIME,USA_SSHS,USA_EYE,USA_RMW,USA_WIND,USA_PRES,USA_ROCI,USA_POCI,USA_PRES) %>%
    ###  there are also two columns that describe the maximum wind speed and its maximum distance in the storm
    ###  this is not provided for each quadrant and so is assumed to be circular and symmetric
    mutate(quad=list(c("NE","SE","SW","NW")),  
           speed=as.numeric(USA_WIND),  
           source="RMW")%>%  
    dplyr::select(-USA_WIND) %>%  
    tidyr::unnest(quad) %>%  
    rename(dist=USA_RMW)) %>%  
    mutate(across(c(dist, speed,USA_EYE,USA_ROCI,USA_POCI,USA_PRES), ~as.numeric(.)),
           USA_SSHS_lab=paste0("Cat. ",USA_SSHS))
```
</details>
<details>
<summary>Fit models from previous hurricane data</summary>
We need these to reconstruct wind extents for any given storm with limited information. Here are non-linear asymptotic models as well as log-tansformed linear models of the wind speed distance as a function of wind speed and storm size by quadrant:
```{r plot models,echo=TRUE,warning=FALSE,message=FALSE}
###  Asymptotic models by storm size and quadrant
ggplot(quaddist%>%filter(!is.na(dist)),aes(x=speed,y=dist,col=factor(quad)))+
  geom_point(size=0.5)+
  geom_point(data=quaddist%>%group_by(quad,speed,USA_SSHS) %>%summarise(dist=mean(dist,na.rm=T)))+
  stat_smooth(method = "nls", formula = y ~ SSasymp(x, Asym, R0, lrc),se=FALSE)+
  facet_wrap(~USA_SSHS)+
  labs(col="Quadrant")+
  ggtitle("North Atlantic Tropical Cyclone Wind Speed Distance by Storm Saffir-Simpson Scale\nAsymptotic models")+
  xlab("Wind Speed (kts)")+ylab("Max. distance from center (nmi)")

####  close up of Category 5 storms
ggplot(quaddist%>%filter(!is.na(dist),USA_SSHS==5),aes(x=speed,y=dist,col=factor(quad)))+
  geom_point(size=0.5)+
  geom_point(data=quaddist%>%group_by(quad,speed,USA_SSHS) %>%summarise(dist=mean(dist,na.rm=T)),size=1)+
  stat_smooth(method = "nls", formula = y ~ SSasymp(x, Asym, R0, lrc),se=FALSE)+
  facet_wrap(~USA_SSHS_lab)+
  labs(col="Quadrant")+
  ggtitle("North Atlantic Tropical Cyclone Wind Speed Distance by Storm Saffir-Simpson Scale\nAsymptotic models")+
  xlab("Wind Speed (kts)")+ylab("Max. distance from center (nmi)")


###  a linear model of the wind speed distance as a function of wind speed and storm size
###  fit doesnt seem as good as the non-linear model
ggplot(quaddist,aes(x=log(speed),y=dist,col=quad))+
geom_point()+
  geom_point(data=quaddist%>%
               filter(!is.na(dist))%>%
               group_by(quad,USA_SSHS,speed) %>%
               summarise(dist=mean(dist,na.rm=T))) +
  stat_smooth(method="lm", formula=y~log(x),se=TRUE)+
  facet_wrap(~USA_SSHS)+
  labs(col="Quadrant")+
  ggtitle("North Atlantic Tropical Cyclone Wind Speed Distance by Storm Saffir-Simpson Scale\nLinear models")+
  xlab("Wind Speed (kts)")+ylab("Max. distance from center (nmi)")
```
In general, the SW quadrant has higher winds closer to the center, and the NE quadrant has the same wind speeds farther out. Storm size also seems to be important. The linear model does not seem to fit as well as the asymptotic models. In either case, models generally only compute for true hurricanes/typhoons with Saffir-Simpson scale larger than 0. We cant easily include an interaction term for an Asymp model, but that is approached in further steps. 

##  Compare models
```{r compare models,include=TRUE}
modquad <- lm(data=quaddist%>%filter(USA_SSHS>=1),dist~log(speed):quad)  
modquad_asymp <- nls(dist ~ SSasymp(speed, Asym,R0, lrc), data = quaddist%>% filter(USA_SSHS>=1),control = list(maxiter = 500))
summary(modquad_asymp)
summary(modquad)
AIC(modquad,modquad_asymp)
BIC(modquad,modquad_asymp)
```
By AIC and BIC, the asymptotic model is marginally better, not enough for a true determination of fitness. Try new models grouped by storm size (saffir simpson and or quad, rather than a single model with these as covariables (in the linear model's case). 

```{r grouped models,echo=TRUE,include=TRUE,warning=FALSE,message=FALSE}
##  create new dataset with all wind speeds present for each storm size and quadrant
quaddist_new <- quaddist %>%
  filter(USA_SSHS>=1)%>%
  distinct(USA_SSHS,quad)%>%
  group_by(USA_SSHS,quad) %>%
  reframe(speed=1:175)%>%
  nest_by(USA_SSHS,quad) 

##  build models on the original data, nested by both size and quadrant
modquads <- quaddist%>% filter(USA_SSHS>=1) %>%
  group_nest(USA_SSHS,quad) %>%
  mutate(data2=quaddist_new$data,  ###  predict fitted models onto the new dataset with the full range of windspeeds
         model_asymp = map(data,~nls(dist ~ SSasymp(speed, Asym,R0, lrc), data = .,control = list(maxiter = 500))),
         model_lm = map(data,~lm(dist ~log(speed), data = .)),
         pred_asymp=map2(model_asymp,data2,predict),
         pred_lm=map2(model_lm,data2,predict)) 
### take the predicted and original values to plot with
modquad_df <- modquads %>%
  select(-c(data,model_asymp,model_lm)) %>%
  tidyr::unnest(c(data2,pred_asymp,pred_lm)) %>%
  left_join(modquads%>% tidyr::unnest(data)%>%select(USA_SSHS,quad,speed,dist),by=c("USA_SSHS","quad","speed")) %>%
  distinct(USA_SSHS,quad,speed,dist,.keep_all=T)

ggplot(modquad_df %>% group_by(USA_SSHS,quad,speed) %>%
         summarise(dist=mean(dist,na.rm=T),
                   speed=mean(speed),
                   fit_lm=mean(pred_lm),
                   fit_asymp=mean(pred_asymp)),
       aes(x=speed,y=dist,col=factor(quad)))+
  geom_point()+
  facet_wrap(~USA_SSHS)+
  geom_line(aes(x=speed,y=fit_asymp,lty="asympt."))+
  geom_line(aes(x=speed,y=fit_lm,lty="linear"))+
  scale_linetype_manual(values=c(1,2))+
  labs(lty="model type",col="Quadrant")+
  ggtitle("North Atlantic Tropical Cyclone Wind Speed Distance by Storm Saffir-Simpson Scale\nGrouped models")+
  xlab("Wind Speed (kts)")+ylab("Max. distance from center (nmi)")

##compare the models
bind_rows(map2(modquads$model_asymp,modquads$model_lm,~AIC(.x,.y)) %>%
                        bind_rows() %>%
                        rename(score=AIC) %>%
                        mutate(USA_SSHS=rep(modquads$USA_SSHS,each=2),
                               quad=rep(modquads$quad,each=2),
                               model=rep(c("asymp","lm"),20),
                               type="AIC"),
                      map2(modquads$model_asymp,modquads$model_lm,~BIC(.x,.y)) %>%
                        bind_rows() %>%
                        rename(score=BIC) %>%
                        mutate(USA_SSHS=rep(modquads$USA_SSHS,each=2),
                               quad=rep(modquads$quad,each=2),
                               model=rep(c("asymp","lm"),20),
                               type="BIC"))%>%
  group_by(USA_SSHS,quad) %>%
  mutate(bestAIC=model[score==min(score[type=="AIC"])],
         bestBIC=model[score==min(score[type=="BIC"])])
```
The asymptotic model is a better fit for most combinations of quadrant and saffir simpson scale. More importantly, the asymptotic model never drops below zero, which is important for predicting high winds near the eye of the storm.

Also need to model the eyewall radius. For these, its important to set the intercept to 0. We know that the eyewall radius can't be negative and it cant be more than the extent of maximum winds, and setting the intercept to 0 assures that.

```{r eyemodels,include=TRUE,warning=FALSE,message=FALSE}
eyemod <- quaddist %>% 
  group_by(SID,ISO_TIME) %>%
  filter(dist==min(dist,na.rm=T),USA_EYE!=" ")%>%
  mutate(USA_EYE=as.numeric(USA_EYE)/2) %>%
  ungroup() %>%
  group_nest(USA_SSHS) %>%
  mutate(model=map(data,~lm(USA_EYE~0 + dist,data=.)),
         pred=map(model, predict)) 


ggplot(data=eyemod %>% tidyr::unnest(c(data,pred)),
       aes(x=dist,y=USA_EYE))+
  geom_point()+
  facet_wrap(~USA_SSHS)+
  stat_smooth(method="lm",formula="y~x+0")+
  geom_line(aes(y=pred),lty=2)+
  geom_abline(slope=1,intercept=0)+
  geom_vline(xintercept=0)+
  geom_hline(yintercept=0)+
  ggtitle("North Atlantic Tropical Cyclone Eye Wall Distance and Distance of Maximum Wind by Storm Saffir-Simpson Scale\nGrouped models")+
  xlab("Minimum Distance of Maximum Wind (nmi)")+ylab("Eyewall Radius (nmi)")
```

These models look decent for "large" storms (2-5) but begin to weaken considerably for "small" storms (0 & 1). We should use them cautiously and scrutinize any predictions made. 

Finally, we need models for the overall storm size or its outer radius. This is important for constraining the interpolations made later. The storm size is coded in IBTrACS as  the radius of last closed isobar, or ROCI. Its pressure is codes as the pressure at the last closed isobar, or POCI.
```{r rocimodels,include=TRUE,warning=FALSE,message=FALSE}
ROCImod <- quaddist %>% 
  group_by(SID,ISO_TIME) %>%
  filter(dist==max(dist,na.rm=T),!is.na(USA_ROCI)) %>%
  ungroup() %>%
  group_nest(USA_SSHS) %>%
  mutate(model=map(data,~lm(USA_ROCI~dist,data=.)),
         pred=map(model, predict))
ggplot(data=ROCImod %>% tidyr::unnest(c(data,pred)),
       aes(x=dist,y=USA_ROCI))+
  geom_point()+
  facet_wrap(~USA_SSHS)+
  stat_smooth(method="lm",formula="y~x")+
  geom_line(aes(y=pred),lty=2)+
  geom_abline(slope=1,intercept=0)+
  ggtitle("North Atlantic Tropical Cyclone Outer Isobar Distance and Maximum Swath Distance\nGrouped models")+
  xlab("Maximum Distance of Minimum Wind (nmi)")+ylab("Radius of Last Closed Isobar (nmi)")
```

For larger storms, the outer radius/diameter can be reasonably predicted from the largest known wind swath. We also need the minimum pressure of the system. This is important for predicting the pressure at any point in the storm, which is imortant for approximating the density of air, which gets us to the force or that air blowing at a particulat speed, which finally gives us the energy dissipated by that mass of air on a surface. 
```{r pocimod,include=TRUE,warning=FALSE,message=FALSE}
POCImod <- quaddist %>% 
  filter(!is.na(USA_POCI),!is.na(USA_PRES),(USA_POCI-USA_PRES)>0) %>%
  group_by(SID,ISO_TIME) %>% 
  mutate(USA_WIND=max(speed),
         Pressdif = USA_POCI-USA_PRES) %>%
  ungroup()%>%
  group_nest(USA_SSHS) %>%
  mutate(model=map(data,~lm(Pressdif~USA_WIND,data=.)),
         pred=map(model, predict))
```
```{r pocimodplot,warning=FALSE,message=FALSE}
ggplot(POCImod %>% tidyr::unnest(c(data,pred)),
       aes(x=USA_WIND,y=Pressdif))+
  geom_point()+
  facet_wrap(~USA_SSHS)+
  stat_smooth(method="lm",formula=y~x)+
  geom_line(aes(y=pred),lty=2)+
  ggtitle("North Atlantic Tropical Cyclone Pressure Difference and Maximum Wind Speed\nGrouped models")+
  xlab("Maximum Wind Speed (kts)")+ylab("Storm Pressure Difference (mbar)")
```

With these models, we can reconstruct a cyclone's wind and pressure fields from limited information. This allows us to predict velocity, pressure, and power dissipation at any point in a storm based on observations from previous storms. In most cases, these models are only reasonable for true hurricanes/typhoons with a Saffir-Simpson rating greater than 0. Save the models to use in later steps. 

```
saveRDS(modquads%>% select(-c(data,pred)),"modquads.rds")
saveRDS(eyemod%>% select(-c(data,pred)),"eyemod.rds")
saveRDS(ROCImod%>% select(-c(data,pred)),"ROCI.rds")
saveRDS(POCImod%>% select(-c(data,pred)),"POCImod.rds")

```

```{r copying previous instance to git, include=FALSE}
   # add this chunk to end of mycode.rmd
   file.copy(from="README.md", overwrite=TRUE,to=paste0(dirname(getwd()),"/README.md"))
   file.copy(from="README_files", overwrite=TRUE,to=paste0(dirname(getwd()),"/README_files"),recursive=TRUE)
```
</details>

<details>
<summary> Reconstruct wind fields </summary>
The IBTrACS data and other products from the National Hurricane Center typically only include information regarding 34,50, and 64 knot sustained winds. Since 2023, the data also includes information about the maximum sustained winds within the storm. For many use cases, this is not enough information and we need to know the estimated winds anywhere in a storm and at any given time. 

Using the models from above, we can now build wind fields for any historical storm using the limited information given.

Pick a storm
```{r loading, include=TRUE,warning=FALSE,message=FALSE}
library(sf)
Maria <- read_sf("C:\\Users\\BenjaminBranoff\\OneDrive - USDA\\Hurricanes\\IBTrACS.NA.list.v04r01.lines.shp")%>%
  filter(SEASON==2017,NAME=="MARIA")
bounds <- read_sf("./Cyclones/data/ne_110m_admin_0_countries.shp")
###  plot the 
plot(st_geometry(Maria%>%st_transform(st_crs(bounds))),col="red")
plot(st_geometry(bounds),add=T)
####  only includes the track
###  use the included wind
swaths <- apply(Maria,MARGIN=1,FUN=function(x) (make_swaths(1,atracks=IBTrACS,mod=modquads,emod=eyemod,
            rocimod=ROCImod,pocimod=POCImod,tracks=x,fillwinds=TRUE)))
lines <- lapply(swaths,"[[",2)
lines <- bind_rows(lines)
plot(lines[lines$kts>100,"kts"],add=TRUE)
```
</details>